An I/O Automata Framework for Distributed System Programming

### 2011-07-13 papers/paper1/paper.tex
Worked on system model.

\documentclass[letterpaper, twocolumn]{article}

@@

\title{Threads Considered Harmful}
\author{Justin R. Wilson \and Christopher D. Gill}
\date{}

@@

\begin{abstract}
Our ability to realize sophisticated distributed systems such as those in cyber-physical systems and pervasive computing is predicated on a solid model and implementation of asynchrony and concurrency.
Threads, the de facto standard in practice, are not a suitable foundation because 1) ad hoc communication among threads prohibits software reuse, 2) threads are difficult to treat formally and therefore \emph{very} difficult to treat informally, and 3) threads are ill-suited for the asynchronous semantics present in the systems we desire to build.
In this paper, we outline the semantics necessary for building sophisticated distributed systems and connect these to the I/O automata formal model.
We then describe an implementation of the I/O automata formal model and cover extensions to the model necessary for operating in a dynamic environment.
Finally, we conclude with an evaluation of our I/O automata framework.
\end{abstract}

@@

One of the forces that prevents the development of any kind of sophisticated software system is accidental complexity~\cite{brooks_nsb}.
In the context of distributed systems, a major source of accidental complexity is a mismatch between the conceptual model of the system and the technology used to implement the system.
The mismatch between the conceptual model and implementation vehicle introduces defects due to translation errors, slows rapid prototyping, and retards iterative refinement.
While there is no ``silver bullet''~\cite{brooks_nsb}, reducing the accidental complexity derived from the mismatch between concept and implementation will increase productivity and make new levels of sophistication possible with existing levels of effort.

@@

Third, threads are ill-suited for the asynchronous semantics present in the systems we desire to build.
The distributed systems we desire to build are asynchronous at both the device and network levels.
Embedded devices must respond to asynchronous events in the environment.
The interrupt-driven TinyOS operating system for wireless sensor network motes is an exemplar in this area.
The fundamental communication primitive in real-world distributed systems is asynchronous message passing.
Realizing asynchrony in a threaded program requires the use of signals (interrupts) and/or (a)synchronous I/O multiplexing.
Signals are primitive, rigid, and difficult to use correctly.
I/O multiplexing, e.g., a select loop, is limited to operating system abstractions, e.g., file descriptors, and induces a reactive state machine structure that is difficult to understand and maintain.

@@

@@

\begin{itemize}
  \item State
  \item Communication
  \item Asynchrony
  \item Concurrency
\end{itemize}

I/O Automata
\begin{itemize}
  \item Compare with UNITY
  \item Compare with esterel
  \item Compare with pi calculus
  \item Compare with Ptolemy
\end{itemize}

>>>>

\documentclass[letterpaper]{article}

@@

\title{An I/O Automata Framework for Distributed System Programming}
\author{Justin R. Wilson \and Christopher D. Gill}
\date{}

@@

\begin{abstract}
In an effort to be portable, system developers often favor existing architectures, languages, operating systems, and standard libraries.
The dominant abstraction in established computing technologies for concurrency and therefore distributed system development is threads.
Consequently, distributed system developers are locked into using threads and experience the difficulties of thread-based development including obscure race conditions, poor support for the semantics of distributed systems, and difficulties integrating software developed using different thread models.
Furthermore, the demand for portability and generality dismisses many of the techniques proposed in research that involve new programming languages, language extensions, or domain-specific solutions.
This paper presents a development framework based on the I/O automata formal model implemented in C++.
The I/O automata model is a proven technique for distributed system design because its semantics match those found in distributed systems.
With the framework, developers can reason about correctness directly from the program text using the I/O automata formalism.
The I/O  automata model also provides a common abstraction for concurrency which facilitates software integration and promotes reuse.
TODO:  What are we doing for evaluation?
%% Our ability to realize sophisticated distributed systems such as those in cyber-physical systems and pervasive computing is predicated on a solid model and implementation of asynchrony and concurrency.
%% Threads, the de facto standard in practice, are not a suitable foundation because 1) ad hoc communication among threads prohibits software reuse, 2) threads are difficult to treat formally and therefore \emph{very} difficult to treat informally, and 3) threads are ill-suited for the asynchronous semantics present in the systems we desire to build.
%% In this paper, we outline the semantics necessary for building sophisticated distributed systems and connect these to the I/O automata formal model.
%% We then describe an implementation of the I/O automata formal model and cover extensions to the model necessary for operating in a dynamic environment.
%% Finally, we conclude with an evaluation of our I/O automata framework.
\end{abstract}

@@

One of the forces that prevents the development of any kind of sophisticated software system is accidental complexity~\cite{brooks_nsb}.
In the context of distributed systems, a major source of accidental complexity is a mismatch between the conceptual model of the system and the technology used to implement the system.
The mismatch between the conceptual model and implementation method introduces defects due to translation errors, slows rapid prototyping, and retards iterative refinement.
While there is no ``silver bullet''~\cite{brooks_nsb}, reducing the accidental complexity derived from the mismatch between concept and implementation will increase productivity and make new levels of sophistication possible with existing levels of effort.

@@

Third, threads are ill-suited for the asynchronous semantics present in the systems we desire to build.
The distributed systems we desire to build are asynchronous at both the device and network levels.
Embedded devices must respond to asynchronous events in the environment.
The interrupt-driven TinyOS operating system for wireless sensor network motes is an exemplar in this area.
The fundamental communication primitive in real-world distributed systems is asynchronous message passing.
Realizing asynchrony in a threaded program requires the use of signals (interrupts) and/or (a)synchronous I/O multiplexing.
Signals are primitive, rigid, and difficult to use correctly.
Synchronous I/O multiplexing, e.g., a select loop, is limited to operating system abstractions, e.g., file descriptors, and induces a reactive state machine structure that is difficult to understand and maintain.
Asynchronous I/O improves efficiency by avoiding the dispatching necessary to determine the I/O operation but are subject to the same limitations of shared-state thread-based programming.

@@

\pagebreak

@@

Our system model is influenced by the semantics of distributed systems, the semantics of devices, and the desire for conceptual integrity.

\paragraph{Local State.}
First, we require that all modules manipulate local state only including modules that might be collocated on the same distributed node. 
Said another way, there is no global or shared state.
Computation designed around the manipulation of local state only is easily modularized and consequently, easier to distribute and execute concurrently.

\paragraph{Message Passing.}
Nodes in distributed systems interact by passing messages over communication links.
Message passing can either be synchronous or asynchronous.
In synchronous message passing, the events of message production and delivery are the same, i.e., atomic.
In asynchronous message passing, the events of message production and delivery are logically distinct events subject to being interposed by other events.
Synchronous message passing is most often used between modules on the same node while asynchronous message passing is most often used between nodes.
Both synchronous and asynchronous message passing are required.
Synchronous message passing is necessary for building composable systems because of it association with atomicity.
Asynchronous message passing is necessary for building real distributed systems as this matches the semantics of the networks found in practice.

\paragraph{Asynchronous Events.}
Events are natural way to deal with the asynchrony involved in receiving messages from the network or interacting with the physical environment.
This implies the model that contains a general treatment of events, i.e., modules can both produce and consume events.
Programs, therefore, are reactive and structured as a set of event produces and consumers.

%% Communicating with external resources such as the operating system or physical environment is foundational to writing useful programs.
%% Consider the history of I/O in the C run-time library.
%% The synchronous function call semantics of the C programming language made it possible to write system calls, e.g., read and write, that transfer control to the operating system which then completes the operation and return control to the process after the system call.
%% This is called \emph{blocking I/O} because the process blocks while it waits for the operating system to complete the I/O.
%% Blocking I/O has the severe restriction that only one I/O source can be used at a time.
%% To remedy this, system calls such as select and poll were introduced that allow a process to monitor I/O sources for events and then respond to these events.
%% Asynchronous I/O was a further advancement that allows for the system (instead of the process) to dispatch a function upon the completion of I/O.
%% Compare this to UNIX signals that are neither general (fixed-number with no associated data) nor principled.
%% Interrupts are difficult to use correctly.

\paragraph{Dynamics.}
The sets of nodes and communication links are dynamic in real distributed systems.
By extension, the number of modules in an individual node and the configuration governing the interaction of modules should also be allowed to vary.
Our system model must support this reality.

\paragraph{Reflection.}
The ability to dynamically configure the set of modules and their interactions implies that modules can know the status of their configuration.
For example, a module might require that a produced event be consumed.

%% \begin{outline}
%% \item State
%%   \begin{outline}
%%   \item There is no shared state in a 
%%   \item Local state only
%%   \item Shared state
%%     \begin{outline}
%%       \item Impossible in distributed systems
%%       \item Dangerous in local systems
%%     \end{outline}
%%   \end{outline}
%% \item Communication
%%   \begin{outline}
%%   \item Atomic asynchronous message passing
%%   \item Network sets size of atom (UDP)
%%   \item Can build reliable streams (TCP)
%%   \item Local equivalent is passing a value
%%   \item Model should lend itself to writing protocols
%%   \end{outline}
%% \item Asynchrony
%%   \begin{outline}
%%     \item Model must have natural support for asynchrony, i.e., event-based
%%     \item Leads to a more efficient implementation because changed state and enabled actions become obvious
%%   \end{outline}
%% \item Concurrency
%%   \begin{outline}
%%     \item Reason about systems using non-deterministic interleaving of atomic actions
%%     \item Model should admit implementations that execute concurrently
%%   \end{outline}
%% \item Dynamics
%%   \begin{outline}
%%     \item Configuration - Edges in graph of communicating components can change at run-time.
%%       \begin{outline}
%%       \item Already required in distributed settings
%%       \item Not addressed in formal models
%%       \end{outline}
%%     \item Extension - Nodes in graph of communicating components can change at run-time.
%%   \end{outline}
%%   \item Reflection
%% \end{outline}

%% I/O Automata
%% \begin{itemize}
%%   \item Compare with UNITY
%%   \item Compare with esterel
%%   \item Compare with pi calculus
%%   \item Compare with Ptolemy
%% \end{itemize}
